Evaluating contact tracing and isolation to control pandemic potential influenza outbreaks

### 2025-04-15 inst/paper/main.tex
update reproduction number and asymptomatic cases info in paper

It has become widely accepted that influenza viruses cannot be controlled with contact tracing and isolation become the rate at which infected individuals become infectious and show symptoms is faster than the latency in the process of contacting contacts (ref). However, with the recent human infections of avian influenza (A/H5N1), mostly among poultry and cattle farmers and the longer estimates of incubation period and serial interval (SI) from historical H5N1 data \citep{Ward2024.12.11.24318702} it may not be the case that contact tracing and isolation is unable to control influenza. Historically, avian influenza (H5N1 and H7N9) have had high severity (case fatality risk $\sim 35\%-55\%$), and therefore sustained transmission in a human population could have substantial mortality and morbidity implications \citep{tannerPandemicPotentialAvian2015}. \\

@@

The reproduction number ($R$) use to simulate transmission for each influenza subtype was 1.1, 1.5 and 2.0. We chose these values and they represent either similar values to previously estimated $R$ values for influenza outbreaks \citep{fergusonStrategiesMitigatingInfluenza2006}, including H1N1 \citep{fraserPandemicPotentialStrain2009}, or they represent an inflated values of current estimates of H5N1 and H7N9 subtypes to evaluate the effectiveness of contact tracing and isolation in the event that pathogen evolution (e.g. recombination of subtypes (ref)) enhances the human-to-human transmissibility of these pathogens. Current estimates for H5N1 and H7N9 subtypes are below unity (i.e. $R < 1$) (\citealt{tannerPandemicPotentialAvian2015}; \citealt{Ward2024.12.11.24318702}, but see \citealt{yangDetectingHumanhumanTransmission2007a} for a H5N1 $R$ estimate above 1) so transmission is expected to stop without intervention (except in a extremely rare case of superspreading event). Therefore evaluating contact tracing and isolation requires simulating outbreaks that are likely to exponentially spread without control measures.

@@

The proportion of asymptomatic infections is assumes to be rare, so we test scenarios with zero asymptomatic tranmission and another with 10\% of cases being asymptomatic. \\

@@

\begin{itemize}
\item This study has looked at the feasibility and effectiveness of contact tracing and isolation for influenza subtypes that have cause epidemic outbreaks.
\item Outbreaks controlled at 50\% of contacts traced and isolated is over ??\%, highlighting the effectiveness of this intervention for infectious diseases are not highly transmissible between humans.
\item Discuss the existing flu vaccines and antivirals that can be used an interventions in addition to NPIs
\item In modelling the effectiveness of contact tracing and isolation we have assumed that there is no capacity limits on contact tracing, which does not hold in large outbreaks such as SARS-CoV-2 (ref). We also ignore potential socio-political factors that would prevent health worker from visiting or contacting infected individuals or contacts of infected individuals (Dhillon and Kelly, 2018). We also did not consider zoonotic spillover infections or formite transmission.
\item Digital tracing has capacity for large-scale contact tracing (pingdemic), also rapid tests are a complementary approach for limiting spread.
\item Antivirals for half the UK, these can be deployed in rapid response to flu pandemic.
\item In this study we have parameterised the transmission model with parameter from avian influenza subtypes (H5N1 and H7N9) and a influenza subtype that caused an epidemic in the recent past (H1N1), and response delays from SARS and COVID-19 outbreaks. However, these results can be used to inform other influenza subtypes that exhibit similar epidemiological characteristics in transmissibility and severity from past outbreaks (H2N2 and H3N2) and still exist in animal reservoirs for potential spillovers back into human populations (ref).
\item This study used a branching process model to simulate individual-level transmission and intervention. The model assumes an infinite population size with no interdependence of contact tracing and isolation between cases whereby a case may be in contact with multiple infectors and only needs to be traced in time by one contact to prevent transmission, as could happened in a fixed random network transmission model. It is unlikely that this model structure difference would have a considerable different on our results for the control of disease outbreaks with contact tracing. We suppose that because our model requires isolation sampled for a single individual, our results are more conservative on the ability to control outbreaks, and that using a random network model would indicate an increased effectivness of contact tracing (Juul and Strogatz, 2023). The infinite population assumption also means that although isolated individuals are assumed to not transmit, there are always other individuals that non-isolated infectors can transmit to, whereas a fixed random network model that assumes optimal indefinite isolation will block of parts of the network and overestimate contact tracing effectiveness (Juul and Strogatz, 2023).
\end{itemize}

>>>>

It has become widely accepted that influenza viruses cannot be controlled with contact tracing and isolation become the rate at which infected individuals become infectious and show symptoms is faster than the latency in the process of contacting contacts (ref). However, with the recent human infections of avian influenza (A/H5N1), mostly among poultry and cattle farmers (Garg et al., 2025) and the longer estimates of incubation period and serial interval (SI) from historical H5N1 data \citep{Ward2024.12.11.24318702} it may not be the case that contact tracing and isolation is unable to control influenza. Historically, avian influenza (H5N1 and H7N9) have had high severity (case fatality risk $\sim 35\%-55\%$), and therefore sustained transmission in a human population could have substantial mortality and morbidity implications \citep{tannerPandemicPotentialAvian2015}. \\

@@

The reproduction number ($R$) use to simulate transmission for each influenza subtype was 1.1, 1.5, 2.5 and 3.5. We chose these values and they represent either similar values to previously estimated $R$ values for influenza outbreaks \citep{fergusonStrategiesMitigatingInfluenza2006}, including H1N1 \citep{fraserPandemicPotentialStrain2009, lesslerOutbreak2009Pandemic2009}, or they represent an inflated values of current estimates of H5N1 and H7N9 subtypes to evaluate the effectiveness of contact tracing and isolation in the event that pathogen evolution (e.g. recombination of subtypes (ref)) enhances the human-to-human transmissibility of these pathogens. Current estimates for H5N1 and H7N9 subtypes are below unity (i.e. $R < 1$) (\citealt{tannerPandemicPotentialAvian2015}; \citealt{Ward2024.12.11.24318702}, but see \citealt{yangDetectingHumanhumanTransmission2007a} for a H5N1 $R$ estimate above 1) so transmission is expected to stop without intervention (except in a extremely rare case of superspreading event). Therefore evaluating contact tracing and isolation requires simulating outbreaks that are likely to exponentially spread without control measures.

@@

The proportion of pandemic-potential influenza infections that are asymptomatic is assumes to be rare. There is evidence from past outbreaks of H1N1 and ??? that some laboratory (polymerase chain reaction) positive cases showed no symptoms or symptoms too mild to be considered influenza-like illness \citep{lesslerOutbreak2009Pandemic2009}. We ran scenarios with zero asymptomatic tranmission and another with 10\% of cases being asymptomatic. \\

@@

\begin{itemize}
\item This study has looked at the feasibility and effectiveness of contact tracing and isolation for influenza subtypes that have cause epidemic outbreaks.
\item Outbreaks controlled at 50\% of contacts traced and isolated is over ??\%, highlighting the effectiveness of this intervention for infectious diseases are not highly transmissible between humans.
\item Discuss the existing flu vaccines and antivirals that can be used an interventions in addition to NPIs
\item In modelling the effectiveness of contact tracing and isolation we have assumed that there is no capacity limits on contact tracing, which does not hold in large outbreaks such as SARS-CoV-2 (ref). We also ignore potential socio-political factors that would prevent health worker from visiting or contacting infected individuals or contacts of infected individuals (Dhillon and Kelly, 2018). We also did not consider zoonotic spillover infections or formite transmission.
\item Digital tracing has capacity for large-scale contact tracing (pingdemic), also rapid tests are a complementary approach for limiting spread.
\item Antivirals for half the UK, these can be deployed in rapid response to flu pandemic.
\item In this study we have parameterised the transmission model with parameter from avian influenza subtypes (H5N1 and H7N9) and a influenza subtype that caused an epidemic in the recent past (H1N1), and response delays from SARS and COVID-19 outbreaks. However, these results can be used to inform other influenza subtypes that exhibit similar epidemiological characteristics in transmissibility and severity from past outbreaks (H2N2 and H3N2) and still exist in animal reservoirs for potential spillovers back into human populations (ref).
\item This study used a branching process model to simulate individual-level transmission and intervention. The model assumes an infinite population size with no interdependence of contact tracing and isolation between cases whereby a case may be in contact with multiple infectors and only needs to be traced in time by one contact to prevent transmission, as could happened in a fixed random network transmission model. It is unlikely that this model structure difference would have a considerable different on our results for the control of disease outbreaks with contact tracing. We suppose that because our model requires isolation sampled for a single individual, our results are more conservative on the ability to control outbreaks, and that using a random network model would indicate an increased effectivness of contact tracing (Juul and Strogatz, 2023). The infinite population assumption also means that although isolated individuals are assumed to not transmit, there are always other individuals that non-isolated infectors can transmit to, whereas a fixed random network model that assumes optimal indefinite isolation will block of parts of the network and overestimate contact tracing effectiveness (Juul and Strogatz, 2023).
\item The reproduction number for H5N1 (including clade 2.3.4.4b) and H7N9 influenza subtypes have been estimated to be far below 1 (Ward et al., Garg et al., 2025). Therefore, the transmissibility scenarios explored in this study are speculative and assume that pathogen evolution increases human-to-human transmission potential. If transmissibility of these avian influenza subtypes remains constant over time then outbreaks are expected to be subcritical and will not produce epidemics and thus not require interventions. If cases of avian influenza are detected with routine surveillance without a known animal source, it may indicate sustained human-to-human transmission and can be a trigger to commence contact tracing and isolation while the number of initial cases is presumed small, which as we show increases epidemic controllability (Figure \ref{fig:prop-outbreak-control-num-init-cases}).
\end{itemize}
